Blood-derived dietary protein promotes sleep in the mosquito Aedes aegypti

Version 1

Abstract

Sleep is a ubiquitous, yet highly variable, behavior across species. The duration and timing of sleep are influenced by ecological demands and dietary context. In the mosquito Aedes aegypti, a blood-feeding insect with specialized nutritional requirements, the relationship between feeding and sleep remains poorly understood. Here, we investigate how blood-derived dietary protein influences sleep regulation. Using postural analysis, videography, and arousal-threshold assays, we establish that immobility bouts of ≥10 minutes reliably define sleep in Ae. aegypti. Infrared activity monitoring revealed that blood-fed females exhibit a marked increase in sleep beginning immediately after feeding and persisting for several days, accompanied by reduced locomotor activity. Notably, this sleep elevation lasts well beyond the cessation of host-seeking behavior, suggesting distinct phases of opportunistic versus determined host pursuit. To determine the dietary basis of this effect, we tested mosquitoes fed a bovine serum albumin (BSA)–based diet. BSA feeding alone was sufficient to mimic the sleep-promoting and hypoactive effects of blood, identifying dietary protein as the key nutritional regulator. Moreover, RNAi-mediated knockdown of the leucokinin receptor (lkr), previously implicated in fluid regulation and feeding, enhanced sleep and reduced activity, implicating LK signaling in postprandial sleep modulation. Together, these findings demonstrate that blood-derived proteins drive sustained increases in sleep and reduced locomotion in Ae. aegypti. This work positions Ae. aegypti as a model for dissecting nutrient-specific regulation of sleep and highlights potential adaptive functions of protein-induced quiescence, such as energy conservation and predator avoidance. More broadly, it provides insight into how specialized diets shape the neural and behavioral architecture of sleep.

Introduction

Sleep duration and timing varies significantly across the animal kingdom [1]. In diverse species, sleep is modified by an animal’s internal and external environment, including social cues, early-life development, stress, and food availability[2,3]. Moreover, evolutionary pressures have shaped species-specific sleep strategies, balancing the restorative benefits of sleep with ecological demands such as predation risk, foraging requirements, and reproductive behaviors [4]. While the evolutionary basis for interspecies variation in sleep remains poorly understood, there is mounting evidence that context-dependent regulation of sleep is adaptive [5,6]. Understanding how these factors interact to regulate sleep not only provides insight into the adaptive significance of sleep but also informs broader questions about the links between sleep, health, and survival across taxa.

Food availability appears to be a potent regulator of sleep [7]. Many species of animals have been shown to forgo sleep to forage under food-deprived conditions [3]. Conversely, flies, fish, and mammals all display post-prandial sleep, demonstrating that sleep is bidirectionally modified by feeding state[8–10]. In insects, both total caloric content and specific dietary nutrients have been shown to influence sleep [11–13]. While this phenomenon is highly conserved, most studies have been performed in Drosophila, a dietary generalist, and far less is known about how diet contributes to sleep regulation in animals that consume more specialized diets [14].

Blood-feeding insects provide a unique opportunity to investigate the effects of diet on behavior [15]. Their obligate dependence on vertebrate blood imposes distinct nutritional constraints that differ markedly from the diets of generalist feeders. In addition, the dramatic physiological changes that occur following a blood meal, such as gut distension and shifts in metabolic state, are likely to influence neural circuits that govern sleep and activity. The mosquito, Aedes aegypti, is a model for understanding the effects of blood feeding on behavior [16,17]. These insects rely exclusively on blood to complete their reproductive cycle, making them an ideal system to study the behavioral consequences of this specialized diet. While sleep has been described in this system, the effects of blood feeding on sleep regulation have not been investigated [18,19].

There is evidence to suggest that blood feeding and dietary protein influence sleep in Ae. aegypti. Following a blood meal, females suppress both host seeking and feeding for several days [20]. Moreover, an artificial protein-based diet composed of γ-globulins, hemoglobin, and albumin is sufficient to stimulate egg production [21], revealing dietary protein as a critical factor in reproduction and in the suppression of feeding. This suggests that dietary proteins, rather than other components of blood, underlie these behavioral shifts. Nevertheless, the specific amino acids that mediate the regulation of reproductive and feeding behaviors remain unidentified. Examining sleep across different feeding conditions therefore provides an opportunity to determine which dietary macronutrients regulate sleep in Ae. aegypti and to test whether these macronutrients overlap with those that regulate other blood-feeding–associated behaviors.

Here, we systematically defined sleep behavior in mosquitoes and tested the effects of blood feeding and dietary protein on sleep. We found that both blood meals and a bovine serum albumin (BSA)-based protein diet increase sleep for several days. Critically, we found that the period of increased sleep is substantially longer than the cessation of host seeking, suggesting the existence of discreet opportunistic vs determined host-seeking phases. These findings establish Ae. aegypti as a model for investigating the effects of diet on sleep regulation and imply that methods to disrupt mosquito host-seeking must consider both of these behavioral systems.

Results

Sleep has previously been defined as inactivity bouts that are greater than two hours based on postural analysis [19]. We sought to further define periods of immobility used to define sleep using multiple established criteria including arousal threshold, posture, and circadian regulation [22]. To identify species-specific postures during the day and night, individual mated female mosquitos were recorded at high resolution. Key postural differences were identified with inactive and active mosquitoes that were consistent with those previously used to define sleep [19]. Shortly following activity, the hind legs extend upward while the proboscis points downward, often contacting the ground revealing a posture associated with waking activity (Fig. 1A–B). Following immobility lasting on average 20 min during the day, and 11 min in the evening, the hind legs are in contact with the tube and the proboscis becomes elevated (Fig. 1C–E). These findings suggest a shorter duration of immobility is associated with sleep than has previously been reported.

Characterization of sleep and arousal threshold in female Ae. aegypti.

(A-D) Representative images of adult female Ae. aegypti either in active state or during sleeping. E) Quantification of sleep onset latency in mosquitoes using postural recordings from an iPhone 16 Pro. Sleep onset latency was significantly longer during the day than at night (ANOVA with Mann-Whitney test, F1,12=7.792, P=0.0013). Individual data points are shown as gray (day) and black (night) dots. (F-I) A modified version of Drosophila Arousal Tracking (DART) system was used to assess sleep responsiveness in mosquitoes. The setup records mosquito movements while simultaneously delivering mechanical stimuli through a digital analog converter (DAC). All measurements were taken from sleeping mosquitos and recorded hourly, beginning at ZT0. (F) Schematic of the vibratory stimulus setup used to assess percent of sleep responsiveness. (G) Arousal threshold measurements in the DART system. The response probability was greater for animals that had been inactive for 1–9 minutes than that had been inactive for (≥10 minutes) One-way ANOVA F4,145=6.817, (1-5) vs. (6-10) P= 0.2460; (1-5) vs. (11-15) P= 0.0051, N=14. (H-I) Individual mosquito trajectories showing increased sleep duration (H, paired t test, P=0.0478) and sleep bout length (I, paired t test, P=0.0041). Error bars represent the standard error of the mean (±SEM). Here and in subsequent figures, asterisks indicate the level of significance: *P<0.05, **P<0.01, ***P<0.001, ****P<0.0001.

To define further the period of immobility associated with sleep in Ae. aegypti we tested mosquitoes in the Drosophila ARousal Threshold (DART) that has been widely used to measure sleep duration and intensity in flies (Fig 1F) [23–25]. The responsiveness to stimuli was significantly reduced at periods of 10 minutes of immobility or longer (Fig 1G). Sensory responsiveness decreased further for bout lengths greater than 15 min, raising the possibility of light and deep sleep (Fig 1G). Together, these findings suggest that bout lengths of 10 min or longer can be used to define sleep.

Using the definition of 10 min of immobility to define sleep, we then sought to determine whether sleep is under circadian regulation. Periods of immobility were quantified using videography from the DART system. Both total sleep duration, and the average length of each bout were greater in the nighttime as compared to the daytime (Fig 1H-I). Therefore, behavioral posture, and arousal threshold, and circadian regulation support a 10 min immobility threshold in Ae. aegypti.

We applied infrared-based activity monitors that have been widely used to measure sleep in Drosophila, as well as Ae. aegypti, to analysis sleep [19,26]. To determine if diet impacts sleep, we compared differences between sugar or blood fed females. Female mosquitoes aged 3-5 days after eclosion were blood-fed with defibrinated sheep blood, and after three days were allowed to lay eggs (Fig 2A). We focused on the period after oviposition as we were primarily interested in behavioral differences induced by blood feeding that were independent of host-seeking, which is restored by 72hr after feeding [27]. Following oviposition, mosquitoes were lightly anesthetized on ice and loaded into custom-building infrared-based activity monitors, like those regularly used to measure sleep and activity in fruit flies. A sleep profile beginning on day 4 following feeding reveals persistently elevated sleep for blood-fed mosquitoes across the day and night periods (Fig 2B and Fig 2C). Analysis over time reveals that sleep was significantly elevated in blood-fed mosquitos for Days 4 and 5, before returning to sugar-fed levels (Fig 2D). We applied a Markov model that determines propensity to remain asleep, which serves as an indicator of sleep depth [28]. On Day 4 following blood feeding, mosquitoes displayed reduced wake propensity and increased sleep propensity (Fig 2E,F). These findings suggest blood feeding induces long-lasting increases in sleep that extend far beyond the window where host responsiveness/seeking is restored.

Effects of blood feeding on sleep patterns in Ae. aegypti.

(A) Schematic representation of the experimental design. Female mosquitoes, aged 3-5 days post-eclosion (dpe), were either sugar-fed or blood-fed. Following a 72-hour period of oviposition, individual animals were loaded to the infrared-based activity monitors (LAM10, diameter=10mm) used for tracking sleep and activity. (B) Sleep profile (minutes per hour) across the 24-hour cycle on Day 4 post-oviposition. Blood-fed mosquitoes (red line) displayed a consistently elevated sleep profile compared to sugar-fed counterparts (gray line). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the daytime and nightime, respectively. (C) Comparison of total sleep of blood-fed mosquitoes (red cycles) on Day 4 post blood meal to sugar-fed controls (gray diamonds). Blood-fed mosquitoes showed a significant increase in total sleep compared to sugar-fed mosquitoes (F1, 30= 135.8, P<0.0001). (D) Total sleep duration across Day 4 to 9 post-blood meal. Blood-fed mosquitoes (red cycles) exhibited significantly increased sleep compared to sugar-fed controls (gray diamonds) on Days 4 and 5 (Day 4: F1, 30= 135.8, P<0.0001; Day 5: F1,30=79.00, P<0.0001). (E-G) Changes of sleep architecture of blood-fed mosquitoes (red cycles) on Day 4 post blood meal compared to sugar-fed controls (gray diamonds). Blood-fed mosquitoes demonstrated a significantly lower waking propensity, P(Wake), compared to sugar-fed mosquitoes (E, F1,30=29.79, P<0.0001). Blood-fed mosquitoes exhibited a significantly higher sleeping propensity, P(Doze), compared to sugar-fed controls on Day 4 post-blood feeding (F, F1,30=64.44, P<0.0001). Error bars represent the standard error of the mean (±SEM). Data points in the bar graphs represent individual animals. Waking activity (activity per waking minute) was reduced on Day 4 post blood meal compared to that of sugar-fed animals, indicating lower movement intensity after blood feeding (G, F1,30=19.98, P<0.0001).

To determine whether diet also influences activity, we normalized activity to account for sleep differences between conditions, as has previously been described in fruit flies [29]. Validated the increased sleep in blood fed mosquitoes with 30min interval analysis (Fig S1A), the waking activity (activity per waking minute) was reduced compared to sugar fed controls, indicating less intensity of movement (Fig 2G). Analysis through nine-days post-blood feeding reveals that waking activity was reduced in blood-fed mosquitoes through Day 5, before returning to normal (Fig S1B-E). Therefore, blood fed mosquitoes are hypoactive for a time course that mirrors the sleep increases [24].

To confirm that the effects observed are not due specifically to recent oviposition only and further define the time-course associated with dietary changes in sleep, we sought to measure sleep immediately following a blood meal. Mosquito behavior was recorded in 6-well tissue culture plates containing 1% agar and 10% sucrose for the first day following bloodfeeding (Fig 3A). Behavior was then tracked using Ethovision as we have previously described in Drosophila [26]. An increase in sleep was detected immediately after blood feeding and this persisted over the 24 hr recording (Fig 3B-C). This was associated with reduced total locomotor activity (3D,E). These findings confirm that blood feeding has both immediate and prolonged effect on sleep and locomotor activity.

Validation of blood feeding induced sleep in female Ae. aegypti using EthoVision.

(A) Schematic of the setup for behavioral recording via EthoVision XT. Female mosquitoes, at the age of 1-3 dpe, were either sugar-fed or blood-fed, and immediately transferred to a 6-well plate after meal for monitoring activities over one LD cycle using EthoVision XT, which allowed for track visualization and analysis of sleep patterns. One animal was housed in one well of the plate. (B) Total daily sleep duration in blood-fed mosquitoes (red) compared to sugar-fed controls (gray). A significant increase of daily sleep exhibited in blood-fed group compared to sugar-fed (F1,12=24.01, P<0.0001). (C) Sleep profiles across the 24-hour cycle for sugar-fed and blood-fed mosquitoes via EthoVision recording. Blood-fed mosquitoes (red) showed a distinct sleep pattern, with increased sleep duration observed during both day and night periods compared to sugar-fed counterparts (gray). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the light and dark, respectively. (D) Representative tracks of sugar-fed (left) and blood-fed (right) mosquitoes under EthoVision recording. The red lines indicate the movement paths of the mosquitoes, illustrating differences in activity levels between the two feeding groups. (E) Total traveled distance (meters per day) by sugar-fed and blood-fed mosquitoes. Blood-fed mosquitoes (red cycles) showed a significantly shorter distance compared to sugar-fed controls (gray diamonds, F1,12=13.06, P=0.0058), indicating reduced activity levels. Error bars represent the standard error of the mean (±SEM) and each dot on the bar graphs represents individual animal.

Protein represents a primary dietary component that contributes to sleep regulation [11,30]. Feeding mosquitoes bovine serum albumin (BSA) has previously been shown induce egg laying indicating that proteins represent the dietary cue in blood that modulates behavior [31]. To determine whether dietary proteins alone increase sleep, we compared sleep in sugar fed mosquitoes and those fed BSA beginning at Day 4 following bloodfeeding using the same time-course applied to quantify the effects of blood feeding on sleep (Fig 4A). Sleep was elevated in BSA-fed mosquitos on Day 4 following feeding compared to sugar-fed controls, and was persistently elevated during the day and the night periods (Fig 4B,C). Sleep measurements over six days revealed that sleep in BSA-fed mosquitos was elevated in BSA-fed mosquitoes for Day 4 and 5, then returned to sugar-fed levels (Fig 4D). The increase in sleep duration was associated with reduced wake propensity and increased sleep propensity supporting deeper sleep (Fig 4E-F). Furthermore, waking activity was reduced in BSA fed animals, suggesting BSA alone is sufficient to induce hypolocomotion (Fig 4G). Together, these findings reveal that the effects of BSA on sleep phenocopy blood feeding, suggesting it is specifically dietary protein that promotes sleep.

Dietary protein increases sleep in female Ae. aegypti via BSA feeding.

(A) Schematic overview of the experimental design. Female mosquitoes, aged 3-5 dpe, were either sugar-fed or fed Bovine Serum Albumin (BSA). Following a 72-hour period of oviposition, sleep was monitored using custom-built infrared activity monitors over a 7-day recording period. (B) Comparison of total sleep of BSA-fed mosquitoes (blue squares) on Day 4 post blood meal to sugar-fed controls (gray diamonds). Total sleep duration of BSA-fed mosquitoes showed a significant increase compared to sugar-fed mosquitoes (F1,30=26.68, P<0.0001). (C) Sleep distribution across one LD 15hr:9hr cycle on Day 4 post-BSA meal. BSA-fed mosquitoes (blue line) demonstrated elevated sleep levels during both day and night periods compared to sugar-fed mosquitoes (gray line). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the daytime and nighttime, respectively. (D) Total sleep duration across Days 4 to 9 post-BSA feeding. BSA-fed mosquitoes (blue) exhibited significantly increased sleep compared to sugar-fed controls (gray) on Days 4 and 5 post-BSA feeding (Day 4: F1,30=26.68, P<0.0001; Day 5: F1,30=15.17, P=0.0005). (E-G) Changes of sleep architecture of BSA-fed mosquitoes (blue) on Day 4 compared to sugar-fed mosquitoes (gray). Significantly lower P(Wake) (E, F1,30=22.90, P<0.0001) and higher P(Doze) (F, F1,30=12.91, P=0.0003) was showed in BSA-fed population compared to sugar-fed group on Day 4 post-BSA feeding. Data points in the bar graphs represent individual animals and error bars represent the standard error of the mean (±SEM). BSA-fed mosquitoes exhibited reduced waking activity compared to sugar-fed controls on Day 4 post-BSA feeding (G, F1,30=23.63, P<0.0001).

The Ae. aegypti leucokinin receptor (lkr) (VectorBase ID: AAEL006636), has been previously shown to modulate fluid excretion following a blood meal and sugar perception [32–38], and has been implicated in regulating sleep in Drosophila [39], raising the possibility that it also impacts sleep in mosquitoes. To test for the role of LK signaling in sleep regulation, we targeted lkr in blood fed animals and measured the effects on sleep. Briefly, mosquitoes were blood fed, then injected with dsRNA targeted to either lkr or eGFP and tested for sleep immediately following injection and oviposition period (Fig 5A). Surprisingly, loss of lkr increased sleep over control mosquitoes injected with dsRNA targeted to eGFP for Days 4 and 5 following blood feeding (Fig 5B-D). This was accompanied by a reduction in wake propensity and an increase in sleep propensity (Fig 5E,F). Furthermore, waking activity was reduced in lkr knockdown mosquitoes compared to dsEGFP injected mosquitoes (Fig 5G). Therefore, silencing lkr promotes sleep and reduces locomotion similarly to blood feeding or BSA feeding. This raises the possibility that blood-feeding inhibits LK signaling to promote sleep.

Role of the leucokinin receptor (lkr) in sleep regulation of female Ae. aegypti.

(A) Schematic overview of the experimental design for sleep monitoring in dsRNA-injected mosquitoes. (B) Daily sleep duration was significantly induced in dsLKR-injected mosquitoes (magenta cross markers) compared to dsEGFP controls (green diamonds) (F1,25=15.89, P=0.0012). (C) Sleep profiles of female mosquitoes injected with either dsLKR or dsEGFP across the 24-hour cycle of Day one post injection. Mosquitoes with LKR knockdown (magenta) displayed a distinct sleep pattern, with increased sleep observed during both day and night periods compared to dsEGFP group (green). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the light and dark periods, respectively. (D) Total sleep duration across Days 1 to 6 post-siRNA injection. Mosquitoes injected with dsLKR (magenta line with cross markers) exhibited significantly increased sleep compared to controls injected with dsEGFP (green line with diamond markers) on Days 1 and 2 (F1,25=15.89, P=0.0012 for Day 1 and F1,25=15.18, P=0.0015 for Day 2). (E-G) Comparisons of P(Wake), P(Doze) and waking activity of mosquitoes injected with dsLKR to dsEGFP controls on Day 1 post injection. After knockdown of LKR, female mosquitoes demonstrated a significantly lower probability of waking compared to dsEGFP controls (E, F1,25=13.19, P=0.0002). In contrast, the sleeping propensity, P(Doze), was significantly induced in dsLKR-injected mosquitoes compared to dsEGFP controls (F, F1,25=9.48, P=0.0012). dsLKR mosquitoes (magenta cross markers) exhibited significantly reduced waking activity compared to dsEGFP controls (green diamonds, G, F1,25=11.71, P=0.0053). Error bars represent the standard error of the mean (±SEM), and unpaired t-test was used for significance: **P<0.01, ***P<0.001, ****P<0.0001.

Discussion

Standardized behavioral characteristics of sleep including elevated arousal threshold, rebound following deprivation and species-specific posture, have been used to characterize sleep in species ranging from the jellyfish Cassiopeia through humans. Sleep can be characterized by physiological changes in brain activity or by monitoring the behavioral correlates that accompany these changes [40,41]. Here, we apply posture analysis, videography, and measurement of arousal threshold to define sleep as periods of immobility lasting greater than 10 minutes. This is longer than the five-minutes of immobility that is widely used to define sleep in Drosophila, and shorter than the 120 minutes that has previously been used to define sleep in Ae. aegypti [19,42]. While the 10-minute definition appears to meet existing criteria, further work could be used to define sleep regulation.

Additional approaches have been used in other models to refine the definition of sleep. For example, indirect calorimetry has been applied in flies and rodents to measure sleep-associated changes in whole-body metabolic rate [43,44]. Application of this approach could define whether the graded changes in arousal threshold observed in mosquitos represent light and deep sleep. In addition, analysis of microbehaviors such as haltere movements associated that are associated with sleep in flies could be applied to mosquitos to better define sleep duration and quality [45,46]. Finally, sleep in flies is associated with changes in local-field potentials and neural activity that could be measured in mosquitos [47,48]. Therefore, while our findings provide support for using 10 minutes of immobility to define sleep, the application of additional methods will further our understanding of sleep quality.

Here, we identify a critical role for dietary protein in regulation of mosquito sleep. A central question is whether specific protein components within blood promote sleep. Prior work in fruit flies has shown that tryptone protein alone is sufficient to induce post prandial sleep [11]. In addition, feeding flies the amino acids D-Serine or D-Glutamine alone increases sleep, suggesting that individual amino acids are sufficient to promote sleep [49]. Studies in mosquitos have attempted to identify dietary components associated with egg production and refeeding. An artificial protein-based diet consisting of gamma-globulins, hemoglobin and albumin was able to stimulate egg production in Ae. aegypti, revealing dietary protein to be a critical factor in reproduction [50–52]. One consistent theme amongst these studies is that blood plasma or BSA can trigger ecdysteriod production and vitellogenesis, but equivalent proportions of amino acids cannot, suggesting diverse amino acids are required. [53,54]. Applying similar approaches of minimal diets and individual amino acid supplementation may help identify wake-promoting dietary amino acids.

Genetic studies in the fruit fly Drosophila melanogaster have led to the identification of highly conserved pathways that regulate both sleep and feeding [40,41,55,56]. Many genes and signaling molecules required for sleep and metabolic regulation are conserved across phyla, including those that regulate the circadian clock, energy stores, and neuropeptide regulation [10,39,57]. For example, Lk or the Lkr are essential for both postprandial sleep and starvation-induced sleep suppression, suggestion the Lk signaling mediates both sleep-feeding interactions [11,39]. Here, we find that targeted knockdown of Lkr increases sleep in amino acid fed flies, suggested a wake-promoting role that may be similar to what has been observed for starvation-induced sleep suppression in flies [39]. Additional analysis of the effects of lkr on sleep across different feeding conditions, as well as the generation of stable mutant loss-of-function will support further dissection of its role in feeding regulation. In addition, the use of dsRNA to identify regulators of sleep in Ae. aegypti provides proof-of-principle for the application of screening approaches in this model that would support the identification of novel sleep regulations.

Our findings raise the possibility that increased sleep and reduced activity following a meal is evolutionary adaptive. There are multiple possible explanations for how these physiological changes might improve survival or reproduction. For example, sleep is proposed to be a mechanism of predator avoidance [58], increasing the likelihood that the female mosquito survives when neither mating nor feeding is required. Second, sleep provides a mechanism of energy conservation that may increase the efficiency of time to the subsequent egg-laying cycle or the likelihood of survival. In the case of Ae. aegypti, we found increased sleep and reduced activity for up to 5 day after blood feeding, long after responsiveness to host cues has been restored. This suggests the existence of an opportunistic phase of host-seeking, whereby a bloodmeal provides sufficient nutrients to not only produce eggs but which allows Ae. aegypti to act more as an ambush predator that can respond to cues and seek hosts when they stray too close to the chosen resting sites without expending much effort/risking predation (Fig 5). Once these resources are exhausted, sleep levels decrease, activity increases in a more determined host seeking mode, where Ae. aegypti may take on greater risks in long-distance host pursuit. Identifying whether this effect is specific to Ae. aegypti or generalizable across other mosquito species and blood-feeding insects will help inform its adaptive function.

Together, these results position Ae. aegypti as a powerful new model for dissecting how diet shapes sleep. By linking blood feeding and dietary protein to sustained changes in sleep and activity, this work establishes a foundation for exploring the molecular and neural mechanisms that couple feeding state to sleep regulation in a highly specialized dietary context. The availability of sophisticated behavioral assays, alongside emerging genetic tools in Ae. aegypti, opens the door to mechanistic studies that can parse the contributions of specific nutrients, signaling pathways, and neuronal circuits to sleep regulation. Looking forward, this system offers an unparalleled opportunity to address broad questions about the adaptive functions of sleep, the evolutionary diversity of sleep–feeding interactions, and the conserved pathways that integrate metabolism and behavior across species.

Materials and methods

Mosquito Husbandry and Maintenance

For all experiments, Aedes aegypti mosquitoes (Liverpool strain, Lvp, originally obtained from Virginia Tech) were cultured at Adelman Lab, Texas A&M University, under controlled conditions. Adult mosquitoes were maintained in mixed-sex groups at a temperature of 26-28°C and 70% (±10%) relative humidity, with a photoperiod of 15 hours of light followed by 9 hours of darkness (LD: 15:9 hr). Eggs were hatched in deionized water; larvae were fed ground Tetramin® fish food daily. Pupae were transferred to mesh-covered cups for adult emergence. For adults, mated females were fed defibrinated sheep blood (Colorado Serum Company, CO) via a Parafilm® membrane feeder at 37°C and allowed for oviposition for 72 hours prior to behavioral experiments. Unless otherwise noted, mated females 3-5 days post-eclosion (dpe) were selected for all experiments.

Sleep, Activity and Arousal Measurement

Analysis of postural changes

The device for recording mosquito posture changes consists of two main components. The first component is an iPhone 16 Pro (Apple Inc., CA, USA), utilized for close-range, high-definition monitoring of posture. The iPhone is securely mounted on a tripod, allowing for easy adjustment of height and distance. The second component is a custom-designed chamber for housing the animals and light sources. This chamber features a compact white exterior with a cut-out front panel, facilitating the placement and adjustment of the iPhone 16 Pro while ensuring clear visibility of the interior. Inside, there is a flat surface made of non-reflective material, serving as a base for placing experimental samples. The subjects are housed in a transparent Pyrex glass tube (length: 100 mm, diameter: 10 mm; TriKinetics Inc., USA), which is vertically fixed to the base. At the upper interior of the device, two adjustable light sources (LD: 15:9 hr) are installed, positioned to uniformly illuminate the sample area. The overall structure is robustly designed to minimize external interference and create a controlled environment for precise measurements and observations. When introducing vibrational mechanical stimuli, the glass tube is placed horizontally on a 3D-printed base, with the monitoring camera positioned directly above it. Unless otherwise specified, mosquitoes aged 3-5 dpe were used for behavioral monitoring following blood feeding. The recorded videos were used for quantification of the length from when mosquitoes, which were in a state of quiescence, initiated putting down the elevated hind leg to the point when the hind leg was fully contact with the substrate. This period is defined as the duration of time required for a mosquito to enter a sleep state.

Activity monitoring system

Sleep and activity were quantified using custom-built infrared activity monitors (LAM10, Trikinetics, Waltham, MA), which allows for intra-red based tracking of behavior. Mosquitoes at the age of 3-5 dpe were briefly anesthetized with ice and loaded individually in to 10mm diameter glass tubes containing sugar food (10% sucrose with 1% agar) following a period of oviposition for 72 hours. Tubes with animals were placed to the monitoring system for recording movement and inactivity, enabling the determination of sleep duration and activity patterns over the subsequent six days. Data were collected as previously described [59], and total sleep duration and activity were then calculated for each mosquito. Sleep was defined as periods of inactivity lasting more than 10 minutes, consistent with the data collected for analysis of postural changes.

Video tracking

Ethovision XT 15 (Noldus Information Technology Inc., VA, USA) system was used for video tracking. In brief, one mosquito was placed in one well of the 6-well plates following blood meal. Videos were recorded for 24 hr (one LD cycle) at 15 frames per second using a USB webcam (LifeCam Studio 1080p HDWebcam, Microsoft) through the video processing software VirtualDub (v1.10.4). To enable recording in darkness and maintain uniform illumination, the cameras’ stock IR-cut filters were removed and swapped for IR long-pass filters (Edmund Optics Worldwide). After acquisition, videos were imported into EthoVision XT 15 to generate x-y coordinates for each mosquito across the entire 24 h recording; subsequently positional data were exported from Ethovision and analyzed using a custom-made Perl script (v5.10.0) and Excel Macro (Microsoft). To acquire the tiny action or movement performed on the same spot, a threshold of 1.8 mm per second was used to compute velocity and locomotor activity metrics as previously described [60].

Arousal threshold

A modified Drosophila Arousal Tracking (DART) system was used for mosquitos [24]. In brief, at 3-5-day-old, sugar-fed female mosquitoes were placed individually in 100 mm glass tubes (Trikinetics, Waltham, MA), which were mounted on a vibration-capable platform. The setup was kept in an environmentally controlled chamber with constant temperature (26-28°C) and humidity (70% ±10%) under a 15:9 h LD cycle. Mosquito behavior was continuously recorded using a high-resolution USB camera (Logitech, 800x600, 5 frames per second) for 24 hours. To measure arousal threshold, vibratory stimuli of increasing intensity [0.6g-1g (acceleration)] were delivered once per hour for 24 hours, beginning at ZT0. Vibrational stimulation, video tracking and analyzing were performed using a custom MATLAB pipeline (MathWorks, Natick, Massachusetts) as described by Faville et al. (2015) [24]. To quantify activity, videos were subsampled at 1 frame per second, and pixel differences between consecutive frames were calculated by subtracting a background reference created from 20 randomly selected frames. Movement was defined as a positional change greater than a specified pixel threshold. Inactivity periods of 10 minutes or more were classified as sleep bouts. For each stimulus, the duration of the mosquito’s ongoing inactivity (bout length) was recorded just before stimulation. A behavioral response was defined as visible movement within 15 seconds after the onset of vibration. The proportion of mosquitoes responding was then calculated for each bout length bin (e.g., 1 min, 3 min, 5 min, etc.).

Sugar, Blood, and BSA Feeding Assay

To assess the effects of different feeding regimens on sleep and activity, female mosquitoes were provided with either a 10% sucrose solution, defibrinated sheep blood (Colorado Serum Company), or Bovine Serum Albumin (BSA, 150 mg/mL in PBS with 1 mM ATP) for 72 hours of feeding. Simultaneously, these females were allowed to acclimate and oviposit for 72 hours prior to the experimental test. For the blood feeding assays, mosquitoes were fed using an artificial membrane feeder (Ref). The BSA feeding was administered by placing the protein solution (150 mg/mL in PBS with 1 mM ATP) in small containers accessible to the mosquitoes. Following feeding, mosquitoes were loaded into activity monitoring system for the subsequent six days.

Double-stranded RNA (dsRNA) synthesis and injection

To knock down AaLKR (AAEL006636), we used the dsRNA sequence against the AaLKR from the Kwon et al. study, which targets a 382 nt region of the transcript [61]. We amplified the template for dsRNA using the primers containing the T7 RNA polymerase promoter on the 5’ ends with Phusion DNA polymerase (NEB) and female adult cDNA as a template. The PCR product (dsRNA template) was integrity-verified by agarose gel electrophoresis and purified using the Nucleospin Gel and PCR cleanup kit (Macherey-Nagel). dsRNA was synthesized using the MEGAscript T7 transcription kit (Thermo Fisher) with 500 ng of template DNA for an overnight reaction at 37 °C. The in-vitro synthesis reaction was treated with RNase-free DNase (TURBO DNase, Thermo Fisher) for 30 min at 37 °C and purified using the MEGAclear transcription cleanup kit (Thermo Fisher). Purified dsRNA was confirmed by agarose gel electrophoresis, quantified by spectrophotometry (NanoDrop One, Thermo Fisher), and stored in aliquots at –80 °C until use. We also made dsRNA against EGFP as a negative control. All primer sequences are found in Table A. 3-5 days old female adult mosquitoes were fed with defibrinated sheep blood (Colorado Serum Company) using an artificial membrane feeder, and only engorged individuals were kept with male mosquitoes (to ensure mating). At about 24 h post blood-feeding, the female mosquitoes were anesthetized on ice and were injected in the thorax with dsRNA (1 μg per individual) using Nanoject II microinjector (Drummond). After the injection, the mosquitoes were maintained on a sucrose solution (30%) for another 48 h, and were provided a cup with damp paper for egg laying. The next day, the mosquitoes were loaded on the DAM instrument.

Table A: primers used to amplify dsRNA templates. Capital letters indicate annealing sequence and lower case letters T7 promoter.

Primer ID 5’-3’

AaKR_T7F taatacgactcactatagggTGTATCAACACGGATCATTGCATGGAATGGC

AaKR_T7R taatacgactcactatagggCATCGCTGCCGTTCAGTGTATTGTTGTTTGC

dsEGFP_F taatacgactcactatagggATGGTGAGCAAGGGCGAGGAGC

dsEGFP_R taatacgactcactat

Statistical Analysis

Comparisons between two groups, either blood feeding or BSA feeding verse sugar feeding controls, were conducted in all cases with a one-way ANOVA followed by Mann-Whitney tests and Welch’s correction where applicable. In the knockdown experiment, sleep duration and architectures were compared between dsRNA targeting lkr and dsEGFP animals. Apart from the data plotted by the paired dot graphs, all the other data were analyzed using the unpaired t-test, and statistical significance was set at P< 0.05. All analyses were performed using InStat software GraphPad Prism (version 10.3.0 for Windows, Boston, Massachusetts, USA), and results are presented as mean ± standard error of the mean (SEM).

eLife Assessment

This important study links blood-derived dietary content to sustained increases in sleep in the mosquito Aedes aegypti. Using multiple independent approaches, the authors provide convincing evidence for blood-induced changes in sleep. These findings have broad implications for understanding how specialized diets regulate sleep across species and for mosquito vector biology.

Reviewer #1 (Public review):

Summary:

The presented investigation aims to expand the sleep definition and its relationship with blood meal and/or circadian clock in the mosquito, Aedes aegypti. The authors exhausted the established sleep analytical paradigm and three behaviour toolkits: LAM10, EthoVision, and DART. They also investigated the potential underlying molecular mechanism by using dsRNA injection (LkR) and a KO mosquito (Cyc-/-).

Strengths:

The authors presented a very solid dataset showing posture changes and an increase in the arousal threshold of the mosquito after 10 minutes of immobility. This is a major clarification and extension to our understanding of insect sleep beyond Drosophila. Inclusion of analytical parameters such as bout length, waking activity and pDoze/Wake provide critical reminder for other investigators of the steps needed for defining sleep in a new species. The investigation, with its technical span in behaviour assays, therefore establishes a good standard for mosquito sleep analysis to the same quality seen in the landmark studies (Shaw et al 2000 and Hendricks et al 2000) for Drosophila sleep. The pioneering data showing a clear effect of blood meal and LkR reduction on locomotion and sleep provides an entry point for further investigations.

Weaknesses:

Despite the versatility of the behaviour and transgenic methods in this manuscript, there are two logical gaps in the conclusion, which are related to the effect of blood meal/BSA/LkR KD on A. aegypti sleep:

(1) Conventionally, a coincidence of sleep increase and locomotion reduction would weaken the certainty of a sleep increase assessment. The authors implied this concurrence observed after blood meal is derived from internal "drowsy" neural state instead of physical "cripple", but they did not use their two high-resolution video tracking velocity or pDoze/Wake to clarify this.

(2) The major molecular component underlying blood meal effect on sleep/locomotion is less certain, because the BSA solution used for feeding contains ATP, which itself is able to enter haemolymph and potentially exerts sleep/locomotion effect. Additionally, the basal or control sleep recording is done after sucrose feeding. It is, however, unclear from the method if this is 10% too? And if the observed sleep level increase after a blood meal is a result of sugar level reduction in the blood (~0.1%).

Reviewer #2 (Public review):

Zhang et al. investigate how blood feeding and dietary protein influence sleep in the mosquito Aedes aegypti. The authors first establish a behavioural definition of sleep using postural analysis and arousal threshold measurements, then demonstrate that both blood meals and a bovine serum albumin (BSA)-based protein diet increase sleep for several days. They further show that RNAi-mediated knockdown of the leucokinin receptor (Lkr) enhances sleep, implicating neuropeptide signalling in the regulation of postprandial sleep. The authors propose that elevated sleep persists well beyond the restoration of host-seeking behaviour, suggesting the existence of distinct "opportunistic" versus "determined" host-seeking phases.

Strengths

The central question is well-motivated, and the experimental approach is systematic. The use of multiple independent methods to characterise sleep - postural analysis, infrared activity monitoring, videography, and arousal threshold - provides converging evidence. The BSA feeding experiment is a particularly effective demonstration that dietary protein, rather than other blood components, is the key regulator of the sleep increase. The conservation of leucokinin signalling in sleep regulation between Drosophila and Ae. aegypti is a noteworthy finding that adds comparative depth.

Weaknesses

(1) Sleep definition.

The authors settle on a 10-minute immobility threshold, but their own data do not convincingly support this choice. The arousal threshold data (Figure 1G) show no significant difference between the 1-5 min and 6-10 min bins (P=0.246), with significance emerging only at the 11-15 min bin. The postural analysis likewise indicates that sleep-associated postures appear at ~20 min during the day and ~11 min at night. A 15-minute threshold would be better supported by the data as presented. The previous literature used 120 minutes for this species (Ajayi et al. 2022), making this a dramatic shift.

(2) Confound of reproduction and sleep.

The primary experimental paradigm measures sleep beginning at Day 4 post-blood feeding, immediately after oviposition. Animals have undergone gut distension, vitellogenesis, and oviposition, and what is being measured as "sleep" could reflect post-reproductive quiescence or recovery rather than diet-induced sleep per se. The BSA experiment partially addresses this, but since BSA also triggers vitellogenesis and egg production (as the authors note), the confound persists.

(3) Opportunistic vs. determined host-seeking hypothesis.

This framework is presented as a key conceptual contribution, but the paper contains no data on host-seeking behaviour. The authors infer two phases from the temporal mismatch between a 72-hour host-seeking suppression window (from prior studies) and elevated sleep through Day 5 (~120 hours). While this is an interesting hypothesis, it requires actual measurement of host-seeking alongside sleep to be substantiated, or at least the caveats need to be discussed more explicitly.

(4) Statistical approach.

The methods describe "one-way ANOVA, followed by Mann-Whitney tests with Welch's correction," which is an internally inconsistent combination: Mann-Whitney is non-parametric and does not use Welch's correction (which applies to t-tests). Throughout the figures, F-statistics (parametric) are reported alongside what appear to be non-parametric tests. The statistical framework needs to be clarified and made consistent. Exact sample sizes per group should also be stated explicitly in the methods for all experiments.

Author response:

The following is the authors’ response to the original reviews.

We thank both reviewers for their thoughtful and constructive evaluations of our manuscript. We are grateful that both reviewers found the study to provide a strong behavioral framework for defining sleep in Aedes aegypti and appreciated the breadth of the behavioral and genetic approaches used. We also appreciate the reviewers’ careful identification of several issues requiring clarification, particularly regarding the interpretation of post-blood-meal sleep, the support for the 10-min sleep threshold, possible nutritional confounds in the BSA experiments, the framing of the host-seeking model, and the description of statistical analyses. In the revised manuscript, we have addressed these concerns by clarifying our rationale, tempering several conclusions, revising the statistical reporting and methods, explicitly stating sample sizes, and expanding the Discussion to better acknowledge limitations and alternative interpretations. Where appropriate, we have also revised the text to distinguish more clearly between increased sleep and reduced locomotion, and to frame mechanistic conclusions more cautiously.

Public Reviews:

Reviewer #1 (Public review):

(1) Conventionally, a coincidence of sleep increase and locomotion reduction would weaken the certainty of a sleep increase assessment. The authors implied this concurrence observed after blood meal is derived from internal "drowsy" neural state instead of physical "cripple", but they did not use their two high-resolution video tracking velocity or pDoze/Wake to clarify this.

Thank you for addressing this point. We understand the need to validate locomotion when used as a readout of sleep. We note that analysis of waking activity is normalized to time spent awake, and therefore should be separate from the time spent inactive that is classified as sleep. Based on the reviewers’ suggestions we have reanalyzed some data and revised the relevant sections in include this analysis.: In brief we performed pDoze/pWake analyses on the two high-resolution tracking video from EthoVision XT system. A velocity threshold of 0.4 mm/s was used, with velocities above 0.4 mm/s defined as wake/activity and velocities below 0.4 mm/s defined as doze/sleep state. pWake and pDoze were defined as proportional time metrics of wake/active (velocity > 0.4 mm/s) and doze/sleep (velocity > 0.4 mm/s) within each LD cycle. The conclusion that sleep is increased following blood feeding is supported by these data. We also note (as described in response to Reviewer 2, that this paper represents a step towards describing sleep in mosquitoes. We hope that future application of approaches used in Drosophila, such as brain imaging and indirect calorimetry will further refine our understanding. Along these lines, we have also included a section in the Discussion about how additional measures, including systems like FlyVista might be applied in the future.

(2) The major molecular component underlying blood meal effect on sleep/locomotion is less certain, because the BSA solution used for feeding contains ATP, which itself is able to enter haemolymph and potentially exerts sleep/locomotion effect. Additionally, the basal or control sleep recording is done after sucrose feeding. It is, however, unclear from the method if this is 10% too? And if the observed sleep level increase after a blood meal is a result of sugar level reduction in the blood (~0.1%).

We thank the reviewer for raising this important issue. We think it is unlikely that the small amount of ATP used for feeding is driving the sleep phenotype. We have now included this point as a caveat within the discussion, and explained its inclusion.

(2) Sucrose concentration in controls

We apologize that this was not clearly stated. Yes, the control mosquitoes were maintained on 10% sucrose, and we have now clarified this explicitly in the Methods and figure legends where relevant.

(3) Could the effect reflect reduced sugar intake rather than blood/protein?

We think this is unlikely, however it cannot be ruled out based on the experiments we have run. We have added discussion of this point. However, we note in fruit flies, this has been studied extensively, and loss of sugar under certain contexts reduces sleep. The points above highlight the need for systematic analysis of the dietary components that contribute to sleep in mosquitoes. While we regret being unable to include them in this manuscript, we note that many of these experiments are challenging (with many controls) and have been ongoing for over a decade (with contributions from many labs) in Drosophila.

Reviewer #2 (Public review):

(1) The authors settle on a 10-minute immobility threshold, but their own data do not convincingly support this choice… A 15-minute threshold would be better supported by the data as presented.

We appreciate this evaluation of the sleep threshold. We chose 10 minutes because the first significance in arousal threshold is at the time-point of 10-15 minutes. Therefore, we believe that sleep bouts longer than 10 minutes should be qualified as sleep. We are particularly interested in why arousal threshold continues to increas at 15 minutes. This is either incomplete sleep between minutes 10 and 15 or the presence of multiple sleep states. We have established a new system in the lab using Zantiks that we believe will allow for simultaneous recording of posture and arousal threshold. We now explicitly comment on this in the discussion, and the need for further analysis of the timeframe for which sleep is defined. Nevertheless, we believe we have honed in on a period of 10-15 minutes that serves as a good proxy for sleep regulation. We hope that this initial description of sleep in mosquitoes provides an initial step towards defining sleep, and that future studies that include techniques applied in Drosophila including brain imaging, indirect calorimetry and additional videography will define more nuanced changes in sleep. We have written in limitations and future opportunities to better define sleep throughout the manuscript.

(2) The primary experimental paradigm measures sleep beginning at Day 4 post-blood feeding, immediately after oviposition... what is being measured as ‘sleep’ could reflect post-reproductive quiescence or recovery rather than diet-induced sleep per se. The BSA experiment partially addresses this, but since BSA also triggers vitellogenesis and egg production, the confound persists.

We agree this is an important concern. Our intent in measuring sleep after oviposition was to isolate prolonged post-feeding effects from the well-established transient suppression of host-seeking that occurs during the first ~72 h after blood feeding. However, as the reviewer notes, this design does not by itself distinguish post-feeding sleep from other physiological processes associated with reproduction, including vitellogenesis, oviposition, or post-reproductive recovery. To address this issue, we included the experiment measuring sleep immediately after blood feeding, before oviposition. We agree, however, that this rationale should have been stated more clearly and that the limitation remains relevant, particularly because BSA can also support egg development. In the revised manuscript, we have therefore: In the current version we have clarified more explicitly that the immediate post-blood-meal recording was included to show that the sleep increase begins before oviposition; We have also tempered our interpretation of the Day 4–5 phenotype to avoid implying that it is purely diet-driven and fully independent of reproductive state; and expanded the Discussion to acknowledge that blood feeding, protein feeding, and reproductive physiology are closely linked in female mosquitoes and that our current experiments do not fully disentangle these processes. These changes frame the data more cautiously: blood/protein feeding is sufficient to induce a sleep-promoting state that begins immediately after feeding and persists into the post-oviposition period, but the relative contributions of nutrient sensing, egg development, and reproductive recovery remain to be determined.

(3) The opportunistic vs. determined host-seeking hypothesis… requires actual measurement of host-seeking alongside sleep to be substantiated, or at least the caveats need to be discussed more explicitly.

We agree with the reviewer. Our intention was to present this as a conceptual model motivated by the temporal dissociation between published host-seeking recovery and the prolonged sleep phenotype observed here, not as a demonstrated behavioral framework directly tested in this study. In the revised manuscript, we have substantially softened this section by clarifying that we did not directly measure host-seeking behavior in the current study; adding explicit caveats that the proposed framework remains speculative until sleep and hostseeking are measured simultaneously in the same animals across the same post-feeding time course. We appreciate this comment and agree that the distinction should be presented as a model for future testing rather than as a central conclusion established by the current data.

(4) The methods describe ‘one-way ANOVA, followed by Mann-Whitney tests with Welch’s correction,’ which is an internally inconsistent combination…

We thank the reviewer for catching this lack of clarity. We apologize for this inconsistency. We have fixed this error. In the revised manuscript, we have carefully rewritten the statistical analysis section to specify: which datasets were analyzed using parametric tests (e.g., ANOVA, with appropriate post hoc comparisons where assumptions were met), which datasets were analyzed using non-parametric tests (e.g., Mann-Whitney), and where Welch’s correction was applied, specifically for unequal-variance t-tests, not Mann-Whitney tests. We have also revised Methods, Figure legends and reporting throughout to ensure that the statistical test named in the text matches the reported test statistics. The changes include statistical methods rewritten for consistency and accuracy, and updated figure legends that include exact sample sizes. In addition, one summary spreadsheet of statistical analysis throughout this study is provided and will be submitted as a supplementary file.

Version 2

Abstract

Sleep is a ubiquitous, yet highly variable, behavior across species. The duration and timing of sleep are influenced by ecological demands and dietary context. In the mosquito Aedes aegypti, a blood-feeding insect with specialized nutritional requirements, the relationship between feeding and sleep remains poorly understood. Here, we investigated how blood-derived dietary protein influences sleep regulation. Using postural analysis, videography, and arousal-threshold assays, we established that immobility bouts of ≥10 minutes reliably define sleep in Ae. aegypti. Mosquitoes lacking the circadian clock gene cycle still maintained daily sleep rhythms but exhibited reduced sleep duration and heightened overall activity. Infrared activity monitoring revealed that blood-fed females showed a marked increase in sleep beginning immediately after feeding and persisting for several days, accompanied by reduced locomotor activity. Notably, this sleep elevation lasted well beyond the cessation of previously reported host-seeking phases, raising the possibility of distinct phases of opportunistic versus targeted host pursuit. To determine the dietary basis of this effect, we tested mosquitoes fed a bovine serum albumin (BSA)–based diet. BSA feeding alone was sufficient to mimic the sleep-promoting and activity-reducing effects of blood, suggesting dietary protein is a major nutritional regulator. Moreover, RNAi-mediated knockdown of the leucokinin receptor (Lkr), which has previously been associated with fluid homeostasis and feeding behavior, resulted in enhanced sleep and reduced activity, implicating mosquito LK signaling in the modulation of postprandial sleep. Together, these findings demonstrate that blood-derived proteins drive sustained increases in sleep and reductions in locomotor activity in Ae. aegypti. This work positions Ae. aegypti as a model for dissecting nutrient-specific regulation of sleep and highlights potential adaptive functions of protein-induced quiescence, such as energy conservation and predator avoidance. More broadly, it provides insight into how specialized diets shape the neural and behavioral architecture of sleep.

Introduction

Sleep duration and timing varies significantly across the animal kingdom [1]. In diverse species, sleep is modified by an animal’s internal and external environment, including social cues, early-life development, stress, and food availability [2,3]. Moreover, evolutionary pressures have shaped species-specific sleep strategies, balancing the restorative benefits of sleep with ecological demands such as predation risk, foraging requirements, and reproductive behaviors [4]. While the mechanistic basis for interspecies variation in sleep remains poorly understood, there is mounting evidence that context-dependent regulation of sleep is adaptive [5,6]. Understanding how evolutionary history and cellular factors interact to regulate sleep not only has potential to provide insight into the adaptive significance of sleep but may also inform broader questions about the links between sleep, health, and survival across taxa.

Food availability potently regulates sleep [7]. Many species of animals have been shown to forgo sleep to forage under food-deprived conditions [3]. In addition, flies, fish, and mammals all display post-prandial sleep, demonstrating that sleep is bidirectionally modified by food consumption [8–10]. In insects, both total caloric content and specific dietary nutrients have been shown to influence sleep [11–13]. While this phenomenon is highly conserved, most studies have been performed in Drosophila, a dietary generalist, and far less is known about how diet contributes to sleep regulation in animals that consume more specialized diets [14].

Blood-feeding insects provide a unique opportunity to investigate the effects of diet on behavior [15]. Their obligate dependence on vertebrate blood imposes distinct nutritional constraints that differ markedly from the diets of generalist feeders. In addition, the dramatic physiological changes that occur following a blood meal, such as gut distension and shifts in metabolic state, are likely to influence neural circuits that govern sleep and activity [16]. The mosquito, Aedes aegypti, is a model for understanding the effects of blood feeding on behavior [17,18]. These insects rely exclusively on blood to complete their reproductive cycle, making them an ideal system to study the behavioral consequences of this specialized diet. While sleep has been described in this system, the effects of blood feeding on sleep regulation have not been investigated [19,20].

There is evidence to suggest that blood feeding and dietary protein influence sleep in Ae. aegypti. Following a blood meal, females suppress both host seeking and feeding for several days [21]. Moreover, an artificial protein-based diet composed of γ-globulins, hemoglobin, and albumin is sufficient to stimulate egg production [22], revealing dietary protein as a critical factor in reproduction and in the suppression of feeding. This suggests that dietary proteins, rather than other components of blood, underlie these behavioral shifts. Nevertheless, the specific amino acids that mediate the regulation of reproductive and feeding behaviors remain unidentified. Examining sleep across different feeding conditions therefore provides an opportunity to determine which dietary macronutrients regulate sleep in Ae. aegypti and to test whether these macronutrients overlap with those that regulate other blood-feeding–associated behaviors.

Here, we examine sleep behavior in Ae. aegypti and tested the effects of blood feeding and dietary protein on sleep. We found that both blood meals and a bovine serum albumin (BSA)-based protein diet increase sleep for several days. Critically, we found that the period of increased sleep is substantially longer than the cessation of host seeking, suggesting the existence of discreet opportunistic vs active host-seeking phases. These findings establish Ae. aegypti as a model for investigating the effects of diet on sleep regulation and imply that methods to disrupt mosquito host-seeking must consider both of these behavioral systems.

Results

Sleep has previously been defined in three mosquito species as inactivity bouts that are greater than two hours based on postural analysis [20]. We sought to further define periods of immobility used to define sleep using multiple established criteria including arousal threshold, posture, and circadian regulation [23]. To identify species-specific postures during the day and night, individual mated female mosquitos were recorded at high resolution. Key postural differences were identified with inactive and active mosquitoes that were consistent with those previously used to define sleep [20]. Shortly following activity, the hind legs extend upward while the proboscis points downward, often contacting the ground revealing a posture associated with waking activity (Figure 1A–B). Following immobility lasting on average 20 min during the day, and 11 min in the evening, the hind legs are in fully contact with the tube and the proboscis becomes elevated (Figure 1C–E). These findings suggest a shorter duration of immobility is associated with sleep than has previously been reported.

Characterization of sleep and arousal threshold in female Ae. aegypti.

(A-D) Representative images of female Ae. aegypti either in active state or during sleeping. These images illustrate postural differences of the hind legs in active or inactive states, which were recorded via videography under the daytime or nighttime. (E) Quantification of averaged time interval that individual animal required to change postures. Averaged time interval to shift postures was significantly longer during the day (clear bar) than that at night (gray-filled bar). Data were assessed for normality using Shapiro-Wilk test. As the Day group violated the normality assumption (P < 0.05), comparisons between Day and Night groups were analyzed using a two-tailed Mann-Whitney U test: P = 0.0021 and U = 174.50. Individual data points are shown as gray or black for selective bouts during daytime (N=16) or black for bouts during nighttime (N=13). (F) Schematic of modified Drosophila ARousal Threshold (DART) system using for assessment of sleep responsiveness in mosquitoes. The mechanical stimuli were delivered once per hour for one LD cycle, beginning at ZT0, with protocol shown in bottom left box. (G) Behavioral threshold assessment in mosquitoes across a single light-dark cycle (LD, 15 h:9 h) using a modified DART system. To determine the behavioral threshold, mosquitoes were subjected to vibratory stimuli of incrementally increasing intensity [0.6g–1.0g (acceleration)] once per hour throughout one complete LD cycle. For each stimulation event, the duration of the ongoing inactive bout prior to stimulation was recorded. Responses were normalized to the mean response within each stimulus intensity bin, such that values represent proportional changes relative to the individual baseline. The data indicated that mosquitoes exhibiting shorter periods of inactivity (1–9 min) frequently responded to the stimulus, whereas those in longer inactive bouts (≥10 min) were significantly less responsive, which suggested the onset of deeper sleep following approximately 10 minutes of continuous inactivity, thereby validating the use of a 10-minute immobility threshold for defining sleep behaviorally in this context. One-way ANOVA F4,140 = 14.93, (1-5 min) vs. (6-10 min) P = 0.2460; (1-5 min) vs. (11-15 min) P = 0.0451. N=29. (H-I) Individual mosquito trajectories showing increased sleep duration (H, paired t test, P = 0.0478) and averaged bout length (ABL) during nighttime (I, paired t test, P = 0.0041). Error bars represent the standard error of the mean (±SEM). Here and in subsequent figures, asterisks indicate the level of significance: *P < 0.05, **P < 0.01, ***P < 0.001.

To define further the period of immobility associated with sleep in Ae. Aegypti, we tested mosquitoes in the Drosophila ARousal Threshold (DART) that has been widely used to measure sleep duration and intensity in flies (Figure 1F) [24–26]. The responsiveness to stimuli was significantly reduced at periods of 10 minutes of immobility or longer (Figure 1G). Sensory responsiveness decreased further for bout lengths greater than 15 min, raising the possibility of light and deep sleep, or that a fraction of the animals is asleep at 10 minutes. Together, these findings suggest that bout lengths of 10 min or longer can be used to define sleep. Using the 10 min of immobility to define sleep, we then sought to determine whether sleep is under circadian regulation. Periods of immobility were quantified using videography obtained in the DART system. Both total sleep duration, and the average length of each sleep bout were greater in the nighttime as compared to the daytime (Figure 1H–I). Therefore, behavioral posture, and arousal threshold, and circadian regulation support a 10 min immobility threshold in Ae. aegypti.

Our findings, and those of others, suggest Ae. aegypti are diurnal [27,28]. The circadian clock influences both sleep duration and the timing of sleep in Drosophila and mouse [29–31]. To determine the effects of the circadian clock in sleep in Ae. aegypti, we compared sleep in mosquitos lacking the core circadian gene cycle (AeCyc−/−) to wildtype (WT) control mosquitoes [32]. We applied infrared-based activity monitors that have been widely used to measure sleep in Drosophila, as well as Ae. aegypti, to analyze sleep [20,33]. Total activity was elevated in AeCyc−/− mosquitoes compared to wildtype during the day and night periods (Figure 2A–B). The dusk activity peak was also largely lost in Cyc−/−mosquitoes, suggesting the circadian clock is required for regulation of daily activity, even under light-dark condition (Figure 2C). While we assessed the loss of Cyc on sleep regulation, sleep was significantly reduced during the day and night periods in Cyc−/− mosquitoes, suggesting that Cyc promotes sleep (Figure 2D–E). Nevertheless, sleep in both wildtype and Cyc−/−mosquitoes was greater during the nighttime than the daytime. We applied a Markov model that determines propensity to remain asleep, which serves as an indicator of sleep depth [34]. The wake propensity was greater in AeCyc−/−while sleep propensity was diminished (Figure 2F–G). Together, these findings suggest that Cyc modulates the timing of locomotor activity and the duration of sleep.

Reduced day and night sleep duration in AeCyc−/− mosquitoes.

(A) Schematic of sleep monitoring in wide type (WT) and AeCyc−/− mosquitoes via infrared-based LAM10 (tube outside diameter = 10 mm, length = 100 mm). (B) Activity profile across day and night cycles (15 h:9 h) for WT (gray) and AeCyc−/−(pink) mosquitoes. AeCyc−/− mutants showed elevated activity during both day and night periods compared to WT animals. Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the daytime and nighttime, respectively. (C) Two-way ANOVA revealed a significant interaction between genotype and ZT time (F1,70 = 36.56, P < 0.0001). WT mosquitoes showed a sharp decrease in activity from ZT15 to ZT16 (paired t-test, P < 0.0001), whereas AeCyc−/− mutants maintained high activity levels at ZT16 (paired t-test, P = 0.4084). At ZT15, AeCyc−/− activity was significantly lower than WT (unpaired t-test, P = 0.0045, Cohen’s d = 1.00), while at ZT16 AeCyc−/− activity was significantly higher than WT (unpaired t-test, P < 0.0001, Cohen’s d = 1.78), suggesting the dusk activity peak was largely lost in AeCyc−/− mosquitoes. (D) Sleep profile across day and night cycles for WT (gray) and AeCyc−/− (pink) mosquitoes. AeCyc−/−mutants showed decreased sleep during both day and night periods compared to WT animals. (E) Two-way ANOVA revealed significant main effects of genotype (F1,70 = 515.66, P < 0.0001) and time of day (F1,70 = 653.43, P < 0.0001), but no significant interaction (F1,70 = 0.98, P = 0.3261). Both WT and AeCyc−/−mosquitoes slept more during the night than during the day (paired t-test, WT: P < 0.0001; AeCyc−/− : P < 0.0001). However, AeCyc−/− mutants exhibited significantly reduced sleep duration compared to WT controls during both the day (Welch’s t-test, P < 0.0001) and the night (unpaired t-test, P < 0.0001). (F) Two-way ANOVA revealed significant main effects of genotype (F1,70 = 16.23, P = 0.0001) and time of day (F1,70 = 92.29, P < 0.0001), with a marginally significant interaction (F1,70 = 17.35, P < 0.0001). Both WT and AeCyc−/− mosquitoes showed higher P(Wake) during the day than at night (WT: Wilcoxon signed-rank test, P < 0.0001; AeCyc−/−: paired t-test, P = 0.0021). Notably, AeCyc−/− mutants exhibited a significantly elevated P(Wake) during the night compared to WT controls (Mann-Whitney U test, U = 41.00, P < 0.0001), whereas daytime P(Wake) did not differ between genotypes (unpaired t-test, P = 0.9081). These results indicated that AeCyc−/− was specifically required for maintaining low wake probability during the night. (G) Two-way ANOVA revealed significant main effects of genotype (F1,70 = 16.80, P < 0.0001) and time of day (F1,70 = 35.69, P < 0.0001), with a significant interaction (F1,70 = 9.38, P = 0.0031). Both WT and AeCyc−/− mosquitoes showed higher P(Doze) during the night than during the day (WT: paired t-test, P < 0.0001; AeCyc−/−: paired t-test, P = 0.0428). AeCyc-/- mutants also exhibited a significantly reduced P(Doze) during the night compared to WT controls (unpaired t-test, P < 0.0001). Error bars represent the standard error of the mean (±SEM). Data points in the bar graphs represent individual animals; N=20 for WT in gray and N=17 for AeCyc−/− in pink.

To determine if diet impacts sleep, we compared differences between sugar or blood fed females. Female mosquitoes aged 3-5 days after eclosion were blood-fed with defibrinated sheep blood, and after three days were allowed to lay eggs (Figure 3A). We focused on the period after oviposition as we were primarily interested in behavioral differences induced by blood feeding that were independent of host-seeking, which is restored by 72 h after feeding [35]. Following oviposition, mosquitoes were lightly anesthetized on ice and loaded into custom-built infrared-based activity monitors, like those regularly used to measure sleep and activity in fruit flies. A sleep profile beginning on day 4 following feeding revealed persistently elevated sleep for blood-fed mosquitoes across the day and night periods (Figure 3B–C). Analysis over time revealed that sleep was significantly elevated in blood-fed mosquitoes for Days 4 and 5, before returning to sugar-fed levels (Figure 3D) [33]. On Day 4 following blood feeding, mosquitoes displayed reduced wake propensity and increased sleep propensity (Figure 3E–F). These findings suggest blood feeding induces long-lasting increases in sleep that extend far beyond the window where host responsiveness/seeking is restored.

Effects of blood feeding on sleep patterns in Ae. aegypti.

(A) Schematic representation of the experimental design. Female mosquitoes, aged 3-5 days post-eclosion (dpe), were either sugar-fed or blood-fed. Following oviposition, individual animals were loaded to the infrared-based activity monitors (LAM10, diameter=10 mm) used for tracking sleep and activity. (B) Sleep profile (min per hour) across the 24 h cycle on Day 4 post-oviposition. Blood-fed mosquitoes (red line, N=31) displayed a consistently elevated sleep profile compared to sugar-fed counterparts (gray line, N=31). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the daytime and nighttime, respectively. (C) Comparison of total sleep of blood-fed mosquitoes (red cycles) on Day 4 post blood meal to sugar-fed controls (gray diamonds). Blood-fed mosquitoes showed a significant increase in total sleep compared to sugar-fed mosquitoes (Unpaired t-test, P < 0.0001). (D) Total sleep duration across Day 4 to 9 post-blood meal. Two-way ANOVA revealed significant effects of time, feeding status, and their interaction (all P < 0.0001). Blood-fed mosquitoes exhibited significantly increased sleep compared to sugar-fed controls on Days 4-6 (Day 4: P < 0.0001; Day 5: P < 0.0001; Day 6: P = 0.021), but not on Days 7-9. (E-G) Changes of sleep architecture of blood-fed mosquitoes (red cycles) on Day 4 post blood meal compared to sugar-fed controls (gray diamonds). Blood-fed mosquitoes demonstrated a significantly lower waking propensity, P(Wake), compared to sugar-fed mosquitoes (E, Mann-Whitney U test = 726, P = 0.0006). Blood-fed mosquitoes exhibited a significantly higher sleeping propensity, P(Doze), compared to sugar-fed controls on Day 4 post-blood feeding (F, Unpaired t-test, P < 0.0001). Error bars represent the standard error of the mean (±SEM). Data points in the bar graphs represent individual animals. Waking activity (activity per waking minute) was reduced on Day 4 post blood meal compared to that of sugar-fed animals, indicating lower movement intensity after blood feeding (G, Unpaired t-test, P = 0.0063).

Effects of blood feeding on sleep in Ae. aegypti females.

(A) Total sleep duration in blood-fed Ae. aegypti versus sugar-fed controls in 30-min time-bin of immobility. Sleep of blood-fed mosquitoes (red cycles) was significantly increased compared to sugar-fed controls (gray diamonds, unpaired t-test, P < 0.0001). (B) Comparisons of the averaged sleep for either sugar-fed or blood-fed individual subjects during daytime and nighttime periods. Two-way ANOVA revealed significant main effects of feeding status (F1,120 = 61.60, P < 0.0001) and time of day (F1,120 = 39.04, P < 0.0001), with no significant interaction (F1,120 = 2.43, P = 0.122). Both sugar-fed and blood-fed mosquitoes slept more during the night than during the day (sugar-fed: paired t-test, P = 0.010; blood-fed: paired t-test, P < 0.0001). Blood-fed mosquitoes showed significantly higher sleep levels during both daytime and nighttime compared to sugar-fed controls (daytime: unpaired t-test, P < 0.0001; nighttime: Mann-Whitney U test, U = 174, P < 0.0001). (C) Daily activity across Days 4 to 9 post-blood meal. Two-way ANOVA revealed significant effects of time, feeding status, and their interaction (all P < 0.0001). Blood-fed mosquitoes (red line) exhibited significantly reduced activity compared to sugar-fed controls (dark gray line) on Days 4 and 5 (Day 4: unpaired t-test, P = 0.0045; Day 5: unpaired t-test, P = 0.0144), but not on Days 6-9 (Day 6: P = 0.337; Day 7: P = 0.304; Day 8: P = 0.431; Day 9: P = 0.431). (D) Total daily activity for sugar-fed and blood-fed mosquitoes on Day 4 post blood meal. Blood-fed mosquitoes displayed significantly lower daily activity levels compared to sugar-fed controls (unpaired t-test, P = 0.0045). (E) Hourly activity profiles across one LD cycle for sugar-fed and blood-fed mosquitoes. Blood-fed mosquitoes showed a distinct reduced activity during both day and night periods. Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the daytime and nighttime, respectively. Each dot represents one individual mosquito, dots in gray indicate sugar-fed mosquitoes (N=31) and red dots are for blood-fed animals (N=31). Error bars represent the standard error of the mean (±SEM), and asterisks indicate the level of significance: *P < 0.05, **P < 0.01, ***P < 0.001.

In Drosophila, periods of sleep for 30 minutes of longer are associated with deep sleep[36]. Using this analysis, blood fed mosquitos slept significantly more than sugar-fed mosquitoes, validating the findings using the previously identified 10-minute-immobility definition (Figure 3–figure supplement 1). Therefore, we sought to reanalyze the effects of blood meals using longer sleep bouts. To determine whether diet induces hypoactivity independent from sleep, we normalized activity to account for differences in sleep between conditions, as has previously been described in fruit flies [33,37]. The waking activity (activity per waking minute) was reduced compared to sugar fed controls, indicating less intensity of movement (Figure 3G). Analysis through nine-days post-blood feeding revealed that waking activity was reduced in blood-fed mosquitoes through Day 5, before returning to normal (Figure 3–figure supplement 1). Therefore, blood fed mosquitoes are hypoactive for a time course that mirrors the sleep increases [24].

To further define the time-course associated with dietary changes in sleep and determine whether changes occur prior to oviposition, we sought to measure sleep immediately following a blood meal. Mosquito behavior was recorded in 6-well tissue culture plates containing 1% agar and 10% sucrose for the first day following blood feeding (Figure 4A), and behavior was then tracked using EthoVision XT system as we have previously described in Drosophila [33]. An increase in sleep and diminished velocity profile was detected immediately after blood feeding and this persisted over the 24 h recording (Figure 4B-C; and Figure 4–figure supplement 1). This was associated with reduced total locomotor activity (Figure 4D-E). These findings confirm that blood feeding has both immediate and prolonged effects on sleep and locomotor activity. To more closely examine the effects of blood feeding on sleep and activity, we sought to extract measures of sleep intensity from video-tracking data. This analysis revealed reduced P(Wake) and increased P(Doze) in blood-fed mosquitoes, confirming that blood feeding increases sleep propensity (Figure 4–figure supplement 1).

Validation of blood feeding-induced sleep in female Aedes aegypti using EthoVision XT.

(A) Schematic of the EthoVision XT-based behavioral recording assay. Female mosquitoes at 1–3 days post-eclosion (dpe) were provided with either a sugar meal or a blood meal and then individually transferred into a 6-well plate immediately after feeding for continuous behavioral monitoring over one LD cycle. EthoVision XT was used for automated locomotor tracking, trajectory visualization, and sleep analysis. (B) Sleep profiles across the 24 h LD cycle in sugar-fed and blood-fed mosquitoes. Blood-fed females exhibited increased sleep relative to sugar-fed controls during both the light and dark phases. Lines represent mean values, and shaded areas indicate ± SEM. White and black bars denote the light and dark phases, respectively. (C) Total daily sleep duration in sugar-fed and blood-fed mosquitoes. Blood-fed females showed a significant increase in total sleep duration compared with sugar-fed controls (unpaired t-test, P < 0.0001). (D) Representative locomotor trajectories of sugar-fed and blood-fed mosquitoes recorded by EthoVision XT. Red lines indicate movement paths, illustrating reduced activity in blood-fed females. (E) Blood feeding significantly reduced total locomotor distance compared to sugar feeding (unpaired t-test, P = 0.015), reflecting the concomitant increase in sleep behavior. Gray denotes sugar-fed mosquitoes (N=13) and red denotes blood-fed mosquitoes (N=17). Error bars represent ± SEM, and each dot in the bar graphs represents one individual mosquito.

Representative velocity traces and characterization of inactivity/sleep bout duration in sugar-fed and blood-fed female Aedes aegypti.

(A) Velocity traces (per frame) of sugar-fed and blood-fed female mosquitoes recorded by EthoVision XT. Gray and red traces represent sugar-fed and blood-fed females, respectively. The dashed horizontal line indicates the velocity threshold of 0.4 mm/s used to classify behavioral states: velocity > 0.4 mm/s was defined as wake/activity, and velocity < 0.4 mm/s as doze/sleep. Shaded regions indicate the dark phase. The upper panel shows a zoomed-in view of representative time windows at higher resolution. (B) Velocity traces (per second) displayed with rolling smoothing over one second to illustrate locomotor events (wake bouts) in sugar-fed and blood-fed female Aedes aegypti. (C) Distribution of wake/active bouts in sugar-fed and blood-fed mosquitoes across one light/dark (LD) cycle. (D) Number of detected wake/active bouts in sugar-fed and blood-fed mosquitoes using different minimum bout duration thresholds (5 s, 10 s, and 30 s). As expected, the number of detected bouts decreased as the minimum bout duration criterion increased. (E) Blood-fed mosquitoes exhibited a significantly lower probability of wakefulness [P(Wake)] compared to sugar-fed controls (unpaired t-test, P < 0.0001), consistent with their increased total sleep duration (Figure 4C). Behavioral state occupancy was quantified based on a velocity threshold of 0.4 mm/s: periods with velocity > 0.4 mm/s were classified as wake/activity, whereas periods with velocity < 0.4 mm/s were classified as doze/sleep. (F) Blood feeding significantly increased the probability of dozing [P(Doze)] compared to sugar feeding (P < 0.0001), inversely mirroring the reduced P(Wake) shown in panel E.

Protein represents a primary dietary component that contributes to sleep regulation [11,38]. Feeding mosquitoes bovine serum albumin (BSA), with ATP included as a phagostimulant, has previously been shown to induce egg laying, indicating that proteins represent a blood-derived dietary cue that modulates behavior [39]. To determine whether proteins are sufficient to increase sleep, we compared sleep in sugar-fed and BSA-fed mosquitoes beginning on Day 4 after feeding, using the same post-feeding time course applied in the blood-feeding experiments (Figure 5A). Sleep was elevated in BSA-fed mosquitoes on Day 4 following feeding compared to sugar-fed controls and was persistently elevated during the day and the night periods (Figure 5B–C). Sleep measurements over six days revealed that sleep in BSA-fed mosquitoes was elevated for Day 4 and 5, then returned to sugar-fed levels (Figure 5D). The increase in sleep duration was associated with reduced wake propensity and increased sleep propensity supporting deeper sleep (Figure 5E–F). Furthermore, waking activity was reduced in BSA-fed animals, suggesting that protein contributes to hyperlocomotion (Figure 5G). Together, these findings reveal that the BSA feeding phenocopies the sleep-promoting and hypoactive effects of blood feeding, suggesting it is specifically dietary protein that promotes sleep.

Dietary protein increases sleep in female Ae. aegypti via BSA feeding.

(A) Schematic overview of the experimental design. Female mosquitoes, aged 3-5 dpe, were either sugar-fed or fed Bovine Serum Albumin (BSA). Following oviposition, sleep was monitored using custom-built infrared activity monitors over a six-day recording period. (B) Comparison of total sleep of BSA-fed mosquitoes (blue squares) on Day 4 post BSA meal to sugar-fed controls (gray diamonds). Total sleep duration of BSA-fed mosquitoes showed a significant increase compared to sugar-fed mosquitoes (Welch’s t-test, P < 0.0001). (C) Sleep distribution across one LD 15 h:9 h cycle on Day 4 post BSA meal. BSA-fed mosquitoes (blue line) demonstrated elevated sleep levels during both day and night periods compared to sugar-fed mosquitoes (gray line). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the daytime and nighttime, respectively. (D) Total sleep duration across Days 4 to 9 post BSA feeding. Two-way ANOVA revealed significant effects of time, feeding status, and their interaction (P = 0.009). BSA-fed mosquitoes (blue) exhibited significantly increased sleep compared to sugar-fed controls (gray) on Days 4 and 5 post BSA feeding (Day 4: Welch’s t-test, P < 0.0001; Day 5: unpaired two-tailed t-test, P = 0.0006), but not on Days 6-9 (Day 6: P = 0.252; Day 7: P = 0.526; Day 8: P = 0.130; Day 9: P = 0.317). (E-G) Changes of sleep architecture of BSA-fed mosquitoes (blue) on Day 4 compared to sugar-fed mosquitoes (gray). Significantly lower P(Wake) (E, unpaired t-test, P < 0.0001) and higher P(Doze) (F, unpaired t-test, P = 0.0003) was showed in BSA-fed population compared to sugar-fed group on Day 4 post-BSA feeding. BSA-fed mosquitoes also exhibited reduced waking activity compared to sugar-fed controls (G, unpaired t-test, P = 0.0082). Each dot in the bar graphs represents one individual mosquito, and blue squares indicate BSA-fed mosquitoes (N=31) and gray diamonds sugar-fed individuals (N=31).

The Ae. aegypti leucokinin receptor (Lkr) (VectorBase ID: AAEL006636), has previously been shown to modulate fluid excretion following a blood meal and sugar perception [40–46], and has been implicated in regulating sleep in Drosophila [47], raising the possibility that it also impacts sleep in mosquitoes. To test for the role of LK signaling in sleep regulation, we targeted Lkr in blood fed animals and measured the effects on sleep. Briefly, mosquitoes were blood fed, then injected with dsRNA targeted to either Lkr or EGFP and tested for sleep immediately following injection and oviposition period (Figure 6A). Surprisingly, loss of Lkr increased sleep over control mosquitoes injected with dsRNA targeted to EGFP for Days 4 and 5 following blood feeding (Figure 6B–D). This was accompanied by a reduction in wake propensity and an increase in sleep propensity (Figure 6E–F). Furthermore, waking activity was reduced in Lkr knockdown mosquitoes compared to control mosquitoes (Figure 6G). Therefore, silencing Lkr promotes sleep and reduces locomotion similarly to blood feeding or BSA feeding. These findings raise the possibility that blood-feeding inhibits LK signaling to promote sleep.

Role of the leucokinin receptor (Lkr) in sleep regulation of female Ae. aegypti.

(A) Schematic overview of the experimental design for sleep monitoring in dsRNA-injected mosquitoes. (B) Sleep profiles of female mosquitoes injected with either dsLKR or dsEGFP on Day 1 post injection. Mosquitoes with Lkr knockdown (pink) displayed a distinct sleep pattern, with increased sleep observed during both daytime and nighttime periods compared to dsEGFP group (green). Light-colored shadows indicate the ±SEM error bar, and white and black boxes indicate the light and dark periods, respectively. (C) Daily sleep duration was significantly induced in dsLKR-injected mosquitoes (pink squares) compared to dsEGFP controls (green diamonds) (unpaired two-tailed t-test, P = 0.0004). (D) Total sleep duration across Days 1 to 6 post-siRNA injection. Mosquitoes injected with dsLKR (red line with pink squares) exhibited significantly increased sleep compared to controls injected with dsEGFP (green line with diamond markers) on Days 1-3 (Day 1: unpaired t-test, , P = 0.0004; Day 2: Welch’s t-test, P = 0.0005; Day 3: Welch’s t-test, P = 0.012), but not on Days 4-6 (Day 4: P = 0.184; Day 5: P = 0.511; Day 6: P = 0.477). (E-G) Comparisons of P(Wake), P(Doze) and waking activity of mosquitoes injected with dsLKR to dsEGFP controls on Day 1 post injection. After knockdown of Lkr, female mosquitoes demonstrated a significantly lower probability of waking compared to dsEGFP injected controls (E, unpaired t-test, P = 0.0011). In contrast, the sleeping propensity, P(Doze), was significantly induced in dsLKR-injected mosquitoes compared to dsEGFP controls (F, unpaired t-test, P = 0.0003). dsLKR injected mosquitoes exhibited significantly reduced waking activity compared to dsEGFP controls (G, unpaired t-test with Welch’s correction, P = 0.0094). Error bars represent the standard error of the mean (±SEM), and asterisks indicate the level of significance: *P < 0.05, **P < 0.01, ***P < 0.001. Each dot represents one individual mosquito, and green diamonds indicate dsEGFP-injected mosquitoes (N=22) and pink squares with cross makers indicate individuals injected with dsLKR (N=22).

Discussion

Standardized behavioral characteristics of sleep including elevated arousal threshold, rebound following deprivation and species-specific posture, have been used to characterize sleep in species ranging from the jellyfish Cassiopeia to humans [4,23,48]. Sleep can be characterized by physiological changes in brain activity or by monitoring the behavioral correlates that accompany these changes [48,49]. Here, we apply posture analysis, videography, and measurement of arousal threshold to define sleep as periods of immobility lasting greater than 10 minutes. This is longer than the five minutes of immobility that is widely used to define sleep in Drosophila, and shorter than the 120 minutes that has previously been used to define sleep in Ae. aegypti [20,50]. While the 10-minute definition appears to meet existing criteria, mosquitos displayed a higher arousal threshold at 15 minutes raising the possibility of multiple sleep states, or that 10 minutes of immobility only results in a subset of animals being asleep. Therefore, additional analysis may be helpful in refining the precise amount of inactivity associated with sleep.

Additional approaches have been used in other models to refine the definition of sleep. For example, indirect calorimetry has been applied in flies and rodents to measure sleep-associated changes in whole-body metabolic rate [51,52]. Application of this approach could define whether the graded changes in arousal threshold observed in mosquitoes represent light and deep sleep. In addition, analysis of microbehaviors such as haltere movements associated that are associated with sleep in flies could be applied to mosquitoes to better define sleep duration and quality [53,54]. Finally, sleep in flies is associated with changes in local-field potentials and neural activity that could be measured in mosquitoes [55,56]. Therefore, while our findings support the use of 10 minutes of immobility to define sleep, the application of additional methods will further improve our understanding of sleep quality. In addition, it would be particularly valuable to apply these analyses to AeCyc−/−mosquitoes with shortened sleep, as well as future genes identified to impact sleep, in order to determine the effects of these genetic mutations on sleep depth and sleep-associated changes in neural activity.

While we have focused on behavioral analysis of sleep, both molecular and physiological changes have been used to define sleep in Drosophila. For example, sleep is associated with a reduction in synaptic proteins throughout the brain, providing a molecular readout of sleep need [57]. In addition, sleep can be identified electrophysiologically through slow-wave oscillations in sleep regulatory networks, through 7-10HZ oscillations in local field potentials, or through wide-spread reductions in network connectivity imaged using Ca2+ sensors [55,58,59]. A number of groups have used GCaMP-expressing mosquitos to measure various aspects of brain activity including chemosensation, and host-seeking behavior [60–62]. The application of these approaches to study sleep could help further define the behavioral and physiological underpinnings of sleep and wakefulness.

Here, we identify a critical role for dietary protein in regulation of mosquito sleep. A central question is whether specific protein components within blood promote sleep. Prior work in fruit flies has shown that tryptone protein alone is sufficient to induce post prandial sleep [11]. In addition, feeding flies the amino acids D-Serine or D-Glutamine alone increases sleep, suggesting that individual amino acids are sufficient to promote sleep [63]. Studies in mosquitoes have attempted to identify dietary components associated with egg production and refeeding. An artificial protein-based diet consisting of gamma-globulins, hemoglobin and albumin was able to stimulate egg production in Ae. aegypti, revealing dietary protein to be a critical factor in reproduction [64–66]. One consistent theme amongst these studies is that blood plasma or BSA can trigger ecdysteriod production and vitellogenesis, but equivalent proportions of free amino acids cannot, suggesting that a diverse complement of amino acids may be required [67,68]. Applying similar approaches of minimal diets and individual amino acid supplementation may help identify the specific dietary components that promote sleep or wakefulness.

A previous study described reduced activity immediately following a blood meal, followed by increased activity post-egg laying [69]. The identified hyperactivity was associated with humidity seeking and dependent on the cycle gene. Conversely, our studies find prolonged reductions in activity following blood-feeding [69]. It is possible that the temporal differences in hypoactivity following feeding are due to the differences in the maintenance of mosquitos prior to, and during the assay. Therefore, further examination of the temporal dynamics associated with hypolocomotion across multiple behavioral paradigms, and particularly those that mirror the animal’s natural ecology will be important.

Genetic studies in the fruit fly Drosophila melanogaster have led to the identification of highly conserved pathways that regulate both sleep and feeding [48,49,70,71]. Many genes and signaling molecules required for sleep and metabolic regulation are conserved across phyla, including those that regulate the circadian clock, energy stores, and neuropeptide regulation [10,47,72]. For example, leucokinin (Lk) and its receptor (Lkr) are essential for both postprandial sleep and starvation-induced sleep suppression, suggestion the Lk signaling mediates both sleep-feeding interactions [11,47]. Here, we found that targeted knockdown of Lkr increases sleep in blood-fed mosquitoes, suggested a wake-promoting role that may be similar to what has been observed for starvation-induced sleep suppression in flies [47]. Additional analysis of the effects of Lkr on sleep across different feeding conditions, as well as the generation of stable loss-of-function mutants, will support further dissection of its role in feeding-dependent sleep regulation. In addition, the use of dsRNA to identify regulators of sleep in Ae. aegypti provides proof-of-principle for the application of screening approaches in this model that would facilitate the identification of novel sleep regulators.

Our findings raise the possibility that increased sleep and reduced activity following a meal is evolutionarily adaptive. Our findings raise the possibility that increased sleep and reduced activity following a meal is evolutionarily adaptive. There are multiple possible explanations for how these physiological changes might improve survival or reproduction. For example, sleep is proposed to be a mechanism of predator avoidance [73], increasing the likelihood that the female mosquito survives when neither mating nor feeding is required. Second, sleep provides a mechanism of energy conservation that may increase the efficiency of time to the subsequent egg-laying cycle or the likelihood of survival. In the case of Ae. aegypti, we found increased sleep and reduced activity for up to 5 days after blood feeding, long after responsiveness to host cues has been restored. This suggests the existence of an opportunistic phase of host-seeking, whereby a bloodmeal provides sufficient nutrients not only for egg production but also to allows Ae. aegypti to act more like an ambush predator that can respond to nearby host cues from a chosen resting site without expending substantial energy or increasing predation risk. Once these resources are exhausted, sleep levels decrease and activity increases, corresponding to a more determined host seeking mode, where Ae. aegypti may take on greater risks in long-distance host pursuit. Determining whether this effect is specific to Ae. aegypti or generalizable across other mosquito species and blood-feeding insects will help inform its adaptive significance.

Together, these results position Ae. aegypti as a powerful new model for dissecting how diet shapes sleep. By linking blood feeding and dietary protein to sustained changes in sleep and activity, this work establishes a foundation for exploring the molecular and neural mechanisms that couple feeding state to sleep regulation in a highly specialized dietary context. The availability of sophisticated behavioral assays, alongside emerging genetic tools in Ae. aegypti, opens the door to mechanistic studies that can parse the contributions of specific nutrients, signaling pathways, and neuronal circuits to sleep regulation. Looking forward, this system offers an unparalleled opportunity to address broader questions about the adaptive functions of sleep, the evolutionary diversity of sleep–feeding interactions, and the conserved pathways that integrate metabolism and behavior across species.

Materials and methods

Mosquito Rearing and Maintenance

All experiments used Aedes aegypti mosquitoes of the Liverpool strain (Lvp), originally obtained from Virginia Tech and subsequently maintained in the Adelman laboratory at Texas A&M University under standardized insectary conditions. The Cyc−/−mutant line was generated as previously described [32] and maintained as a stable homozygous stock in the same facility. Adult mosquitoes were housed in mixed-sex cages at 26–28°C and 70% ± 10% relative humidity under a 15 h light: 9 h dark photoperiod (15 h: 9 h LD). Eggs were surface-sterilized with 2% sodium hypochlorite for 2 min, rinsed thoroughly with deionized water, and hatched in deionized water. Larvae were reared in distilled water and fed daily with finely ground Tetramin® fish food at a standardized, density-dependent regimen. Pupae were individually transferred to mesh-covered emergence cups to ensure accurate tracking of eclosion timing. For blood feeding, mated adult females were offered defibrinated sheep blood (Colorado Serum Company, Denver, CO, USA) through a Parafilm® membrane feeder maintained at 37°C; engorgement was visually confirmed before returning mosquitoes to cages. After blood feeding, females were provided with oviposition substrate (seed germination paper strips in water-filled ovitraps) and allowed a standardized 72-h oviposition period when required. Unless otherwise specified, all experimental assays used mated females aged 3–5 days post-eclosion (dpe).

Sleep, Activity, and Arousal Measurements

Analysis of Postural Changes

Mosquito postural changes were recorded using a two-component imaging system: the first component was an iPhone 16 Pro (Apple Inc., Cupertino, CA, USA), mounted on an adjustable tripod to permit precise control of imaging height and distance; the second component was a custom-built behavioral chamber designed to house animals under controlled light/dark and temperature conditions. The chamber had a white exterior with a cut-out front panel to allow stable camera positioning and unobstructed interior imaging. A flat, non-reflective platform served as the base for sample placement. Mosquitoes were housed individually in transparent Pyrex glass tubes (100 mm in length, 10 mm in diameter; TriKinetics Inc., Waltham, MA, USA), which were secured vertically during standard recordings. Adjustable light sources were mounted in the upper interior of the chamber to provide uniform illumination under a 15 h: 9 h LD cycle. This design minimized external disturbance and maintained a stable environment for behavioral measurements. For experiments involving vibrational stimulation, the glass tube was positioned horizontally on a 3D-printed support platform, with the camera aligned perpendicularly above the sample. Video recordings were analyzed to quantify the interval time between the onset of hind-leg lowering, marking the transition from quiescence, and full contact of the hind leg with the substrate. This interval was operationally defined as the time required to enter a sleep-like state.

Evaluation of Arousal threshold using custom-designed DART system

A modified Drosophila Arousal Tracking (DART) system was used for mosquito videography [25]. Briefly, sugar-fed females 3–5 dpe were individually placed in transparent Pyrex glass 100 mm tubes (outside diameter 10 mm and length 100mm; TriKinetics, Waltham, MA, USA) mounted on a vibration-capable platform. The setup was housed in an environmentally controlled chamber as described above for postural analysis. Mosquito behavior was recorded continuously for 24 h using a high-resolution USB camera (Logitech; 800 × 600 resolution; 5 frames/s). To measure arousal threshold, vibratory stimuli of increasing intensity [0.6–1.0 g (acceleration)] were delivered once per hour over one 15 h: 9 h LD cycle, beginning at Zeitgeber time 0 (ZT0). Stimulation, video tracking, and data analysis were performed using a custom MATLAB pipeline as previous described [25]. For activity quantification, videos were subsampled at 1 frame/s, and pixel differences between consecutive frames were calculated relative to a background reference generated from 20 randomly selected frames. Movement was defined as a positional change exceeding a predefined pixel threshold. Immobility periods lasting ≥10 min were classified as sleep bouts. For each stimulus, the duration of ongoing inactivity (i.e., bout length) was recorded immediately before stimulation. A behavioral response was defined as observable movement within 15 s of vibration onset. The proportion of responding mosquitoes was then calculated for each bout-length bin (e.g., 1 min, 3 min, 5 min).

Activity Monitoring System

Sleep and locomotor activity were quantified using custom-built infrared activity monitors (LAM10, Trikinetics, Waltham, MA, USA), which detect movement via infrared beam breaks across 3-board stacks. Prior to loading, females were assigned to sugar-fed, blood-fed, or bovine serum albumin (BSA)-fed groups as described below. In brief, mosquitoes were anesthetized with ice and individually loaded into 10-mm-diameter Pyrex glass tubes containing a solidified sugar diet (10% sucrose with 1% agar). Tubes were placed horizontally in the monitoring racks, and beam-break events of each board stack were recorded continuously 6–8 days under controlled environmental conditions. Activity counts were binned into 1-min intervals and beam breaks across 3-board stacks on the same monitor were integrated prior to sleep analysis. Immobility was defined as 10 consecutive minutes without beam breaks, which is consistent with the data collected in analysis of postural changes. Total sleep duration, mean activity level, bout number and sleep duration, and other sleep architecture parameters were computed per individual using custom Python scripts (Supplementary file 1).

Video tracking with EthoVision

Video tracking was performed using EthoVision XT 15 (Noldus Information Technology Inc., VA, USA). Individual mosquitoes were transferred into separate wells of a 6-well plate immediately after feeding as previously described in Drosophila [33]. Videos were recorded continuously for 24 h (one LD cycle) at 15 frames/s using a USB webcam (LifeCam Studio 1080p HD Webcam, Microsoft). Video acquisition was performed in VirtualDub (v1.10.4). To permit recording during the dark phase while maintaining uniform illumination, the built-in IR-cut filter of the camera was removed and replaced with an infrared long-pass filter (Edmund Optics Worldwide). After acquisition, videos were imported into EthoVision XT 15 to extract x-y coordinate data for each mosquito across the full 24 h recording period. The instantaneous velocity of each tracked subject was calculated and recorded for every frame. Position and velocity data were exported and further analyzed using a custom Perl script (v5.10.0) and Microsoft Excel macros. To classify behavioral state, a velocity threshold of 0.4 mm/s was used, with velocities above 0.4 mm/s defined as wake/activity and velocities below 0.4 mm/s defined as doze/sleep state.

Sugar, blood, and BSA feeding assay

To assess the effects of feeding status on sleep and locomotor activity, female mosquitoes were fed either defibrinated sheep blood (Colorado Serum Company) or a bovine serum albumin solution (150 mg/mL BSA in PBS supplemented with 1 mM ATP, included as a phagostimulant) using an artificial membrane feeder, and allowed to oviposit for 72 h after feeding when required. Sugar-fed controls were maintained on 10% sucrose and subjected to the same time course as animals provided with blood or BSA. Females that had oviposited were then transferred immediately to the activity monitoring system in 10-mm-diameter Pyrex glass tubes containing a solidified sugar diet (10% sucrose + 1% agar). For EthoVision XT-based assays, mosquitoes were loaded into the tracking system immediately after blood feeding. In all cases, only fully engorged individuals fed with blood or BSA were used for behavioral test, and both experimental and control mosquitoes were briefly anesthetized on ice before loading.

Double-stranded RNA synthesis and injection

To knock down AaLKR (AAEL006636), we used the dsRNA sequence against AaLKR reported by Kwon et al., targeting a 382 nt region of the transcript [75]. The dsRNA template was amplified by PCR from adult female cDNA using Phusion DNA polymerase (NEB) and primers containing the T7 RNA polymerase promoter sequence at the 5′ ends. PCR products were verified by agarose gel electrophoresis and purified using the NucleoSpin Gel and PCR Clean-up Kit (Macherey-Nagel). dsRNA was synthesized in vitro using the MEGAscript T7 Transcription Kit (Thermo Fisher) with 500 ng template DNA in an overnight reaction at 37°C. The transcription reaction was treated with RNase-free DNase (TURBO DNase, Thermo Fisher) for 30 min at 37°C and purified using the MEGAclear Transcription Cleanup Kit (Thermo Fisher). Purified dsRNA was verified by agarose gel electrophoresis, quantified by spectrophotometry (NanoDrop One, Thermo Fisher), and stored in aliquots at −80°C until use. dsRNA against EGFP was generated as a negative control. Primer sequences are listed in Table A. Female mosquitoes 3–5 dpe were fed defibrinated sheep blood (Colorado Serum Company) using an artificial membrane feeder, and only fully engorged individuals were retained with males to ensure mating. Approximately 24 h after the blood meal, females were anesthetized on ice and injected into the thorax with dsRNA (1 μg per individual) using a Nanoject II microinjector (Drummond). After injection, mosquitoes were maintained on 30% sucrose for an additional 48 h and provided with a cup lined with damp paper for egg laying. The following day, mosquitoes were loaded into the behavioral monitoring systems.

Statistical analysis

All statistical analyses were performed in GraphPad Prism (version 10.3.0; GraphPad Software, Boston, MA, USA). Data are presented as mean ± SEM unless otherwise indicated. Sample sizes are provided in the figure legends and/or in the Methods. Data distribution and variance were assessed where appropriate before statistical testing. For two-group comparisons, including knockdown experiments comparing mosquitoes injected with dsRNA targeting AaLkr or dsEGFP, data meeting parametric assumptions were analyzed using unpaired two-tailed t tests, with Welch’s correction applied when variances were unequal. Data that did not meet parametric assumptions were analyzed using Mann–Whitney U tests. For comparisons involving multiple groups, including blood-fed or BSA-fed mosquitoes relative to sugar-fed controls, parametric data were analyzed using one-way ANOVA followed by appropriate multiple-comparison tests, whereas nonparametric data were analyzed using Kruskal–Wallis tests with corrected post hoc comparisons. Paired data shown in paired dot plots were analyzed using paired statistical tests as appropriate. Differences were considered significant at P < 0.05. The statistical test used for each analysis is indicated in the corresponding figure legend and supplementary file (Supplementary file 2).